---
title: "Nurse suicides from ONS data on suicides by occupation up to 2019"
author: "Michelle Jamieson"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load packages

```{r, warning=FALSE, message=FALSE}
library(tidyverse)  # for tidy workflow
library(readxl)  # for reading excel files
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(curl)  # for downloading files
library(glue)  # for programming strings
```


# Load data

## Download data

The ONS publishes a summary of suicides categorised by occupation, as an "ad-hoc" analysis: [ONS website](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/adhocs/10807suicidebyoccupationenglandandwales2011to2018registrations)

Two spreadsheets are provided for England & Wales:

* https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/adhocs/10807suicidebyoccupationenglandandwales2011to2018registrations/1.suicidebyoccupationengland.xlsx
* https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/adhocs/10807suicidebyoccupationenglandandwales2011to2018registrations/2.suicidebyoccupationwales.xls.xlsx


```{r}
url_england_2019 <- "https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/adhocs/10807suicidebyoccupationenglandandwales2011to2018registrations/1.suicidebyoccupationengland.xlsx"
  
url_wales_2019 <- "https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/adhocs/10807suicidebyoccupationenglandandwales2011to2018registrations/2.suicidebyoccupationwales.xls.xlsx"

if (!dir.exists("./data")) dir.create("./data")  # create data dir if doesn't exist

path_england_2019 <- "./data/1.suicidebyoccupationengland.xlsx"
path_wales_2019 <- "./data/2.suicidebyoccupationwales.xls.xlsx"

curl_download(url = url_england_2019, destfile = path_england_2019)
curl_download(url = url_wales_2019, destfile = path_wales_2019)
```


## Clean & import

Occupation classification used is Standard Occupational Classification (SOC).

Both ONS spreadsheets have several tables. Tables 1-4 use increasingly detailed classifications of occupations:

1. Major group
2. Sub-major group
3. Minor group
4. Unit group


```{r}
classification_levels <- c("major","submajor","minor","unit")  # these are used as names for the data tables
cell_ranges <- c(  # these correspond to where in the spreadsheets the data are stored, and correspond to the classification levels above
  "Table 1!B3:U14",
  "Table 2!B3:U29",
  "Table 3!B14:U105",
  "Table 4!B14:U384"
)

data_england_2019 <- map(
  .x = cell_ranges,
  .f = 
    ~read_excel(path = path_england_2019, trim_ws = TRUE, range =.x) %>%
      set_names(x = ., nm = c(names(.)[1:2], paste0("Males_",2011:2019), paste0("Females_",2011:2019))) %>%  # rename columns to include both sex & year
      clean_names() %>%  # lowercase names, no spaces
      slice(-1) %>%  # remove first row, it's used to specify year
      pivot_longer(cols = males_2011:females_2019, names_to = c("sex","year"), names_sep = "_", values_to = "n")  # convert ot tidy format
  ) %>% 
  set_names(x = ., nm = classification_levels)  # name each element of list according to classification level

data_wales_2019 <- map(
  .x = cell_ranges,
  .f = 
    ~read_excel(path = path_wales_2019, trim_ws = TRUE, range =.x) %>%
      set_names(x = ., nm = c(names(.)[1:2], paste0("Males_",2011:2019), paste0("Females_",2011:2019))) %>%  # rename columns to include both sex & year
      clean_names() %>%  # lowercase names, no spaces
      slice(-1) %>%  # remove first row, it's used to specify year
      pivot_longer(cols = males_2011:females_2019, names_to = c("sex","year"), names_sep = "_", values_to = "n")  # convert ot tidy format
  ) %>% 
  set_names(x = ., nm = classification_levels)  # name each element of list according to classification level

## combine England & Wales in a single dataframe
data_2019 <-
  map2(
  .x = data_england_2019,
  .y = data_wales_2019,
  .f = function(x,y) {
    full_join(
      x %>% rename(n_eng = n),
      y %>% rename(n_wal = n),
    ) %>%
      mutate(n_engwal = n_eng + n_wal) %>%
      pivot_longer(cols = c(n_eng, n_wal, n_engwal), names_to = "country", values_to = "n") %>%
      mutate(country = case_when(
        country == "n_eng" ~ "England",
        country == "n_wal" ~ "Wales",
        country == "n_engwal" ~ "England & Wales",
        TRUE ~ NA_character_
      )) %>%
      pivot_wider(names_from = sex, values_from = n) %>%
      mutate(all = males + females) %>%
      pivot_longer(cols = c(males, females, all), names_to = "sex", values_to = "n") %>%
      pivot_wider(names_from = year, values_from = n) %>%
      mutate(all = rowSums(select(.,as.character(2011:2019)))) %>%
      pivot_longer(cols = c("all",as.character(2011:2019)), names_to = "year", values_to = "n")
  }
)
```


# Defining nurses & midwives in data

First I check through potentially relevant codes in SOC2010 across the classification levels, using keywords `nur`, `midw`, and `car`:

```{r}
soc_codes <-
  data_2019 %>% map_df(
    .x = .,
    .f = ~ select(.data = .x, 1:2) %>% rename_all(.funs = ~ c("code", "description")) %>% distinct
  )

soc_codes %>%
  filter(str_detect(description, regex("nurs|midw|car", ignore_case = TRUE))) %>%
  print(n=nrow(.))
```


We define nurses as 3 digit (minor group) categories:

* 223 Nursing and Midwifery Professionals

Or separately in 4 digit (unit group) categories:

* 2231 Nurses
* 2232 Midwives
* 6141 Nursing auxiliaries and assistants


# Analysis

## Number of nurses & Midwives who died by suicide

```{r}
data_2019$minor %>%
  filter(soc_minor_group %in% c(223)) %>%
  filter(country == "England & Wales") %>%
  pivot_wider(names_from = year, values_from = n)
```

### Total suicides by year, country, sex

```{r}
data_2019$minor %>%
  group_by(country, sex, year) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = n)
```


## Top 10 occupations by suicide

```{r}
data_2019$minor %>%
  filter(country == "England & Wales") %>%
  filter(sex == "all") %>%
  filter(year == "all") %>%
  top_n(n = 10, wt = n) %>%
  arrange(desc(n))

data_2019$minor %>%
  filter(country == "England & Wales") %>%
  filter(sex == "females") %>%
  filter(year == "all") %>%
  top_n(n = 10, wt = n) %>%
  arrange(desc(n))

data_2019$minor %>%
  filter(country == "England & Wales") %>%
  filter(sex == "males") %>%
  filter(year == "all") %>%
  top_n(n = 10, wt = n) %>%
  arrange(desc(n))
```


## Ranked Nurses & Midwives

```{r}
data_2019$minor %>%
  filter(country == "England & Wales") %>%
  filter(sex == "all") %>%
  group_by(year) %>%
  mutate(rank = rank(-n, ties.method = "max")) %>%  # rank(-n, ties.method = "max") orders in descending order!
  ungroup %>%
  filter(soc_minor_group %in% c(223))

data_2019$minor %>%
  filter(country == "England & Wales") %>%
  filter(sex == "females") %>%
  group_by(year) %>%
  mutate(rank = rank(-n, ties.method = "max")) %>%  # rank(-n, ties.method = "max") orders in descending order!
  ungroup %>%
  filter(soc_minor_group %in% c(223))

data_2019$minor %>%
  filter(country == "England & Wales") %>%
  filter(sex == "males") %>%
  group_by(year) %>%
  mutate(rank = rank(-n, ties.method = "max")) %>%  # rank(-n, ties.method = "max") orders in descending order!
  ungroup %>%
  filter(soc_minor_group %in% c(223))
```


### All year rankings in 4-digit SOC

```{r}
data_2019$unit %>%
  filter(country == "England & Wales") %>%
  filter(year == "all") %>%
  group_by(sex) %>%
  mutate(rank = rank(-n, ties.method = "max")) %>%  # rank(-n, ties.method = "max") orders in descending order!
  ungroup %>%
  filter(str_detect(description, regex("^nurses|nursing", ignore_case = TRUE)))
```

# Visualization

## Time Series Analysis

```{r}
# Ensure you are using the 'minor' data frame from your list
data_2019$minor %>%
  filter(soc_minor_group %in% c(223)) %>%
  filter(country == "England & Wales") %>%
  ggplot(aes(x = as.integer(year), y = n, group = sex, color = sex)) +
  geom_line() +
  labs(title = "Trend of Suicides Among Nurses and Midwives (2011-2019)",
       x = "Year",
       y = "Number of Suicides",
       color = "Sex") +
  theme_minimal()

```

## Regions and Gender

```{r}
# Use the 'minor' data frame from your data_2019 list
data_2019$minor %>%
  filter(soc_minor_group %in% c(223)) %>%  # Ensure you are focusing on nurses and midwives
  ggplot(aes(x = country, y = n, fill = sex)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Comparison of Suicides by Region and Sex",
       x = "Region",
       y = "Number of Suicides",
       fill = "Sex") +
  theme_minimal()

```

# Top 10 Occupations by Suicide

```{r}
top_occupations <- data_2019$minor %>%
  filter(country == "England & Wales") %>%
  filter(sex == "all") %>%
  filter(year == "all") %>%
  top_n(n = 10, wt = n) %>%
  arrange(desc(n))

ggplot(top_occupations, aes(x = reorder(description, -n), y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Occupations by Suicide Rate",
       x = "Occupation",
       y = "Number of Suicides") +
  theme_minimal()

```

# Ranked Nurses & Midwives by Year

```{r}
library(ggplot2)

plot <- data_2019$minor %>%
  filter(soc_minor_group %in% c(223), country == "England & Wales") %>%  # Filter to include only nurses and midwives
  group_by(year, sex) %>%
  summarise(n = sum(n), .groups = "drop") %>%  # Summarizing total suicides by year and sex
  ggplot(aes(x = as.integer(year), y = n, fill = sex)) +
  geom_area(position = 'stack', alpha = 0.6) +
  labs(title = "Suicides Among Nurses and Midwives by Year and Sex",
       x = "Year",
       y = "Number of Suicides",
       fill = "Sex") +
  theme_minimal()

print(plot)

library(ggplot2)

# Assuming data_2019$minor has been properly prepared with columns year, sex, and n (number of suicides)
ggplot(data_2019$minor, aes(x = as.integer(year), y = n, color = sex)) +
  geom_line() +
  facet_wrap(~country) +  # Creates a separate plot for England and Wales
  labs(title = "Trend of Suicides Among Nurses and Midwives by Sex",
       x = "Year",
       y = "Number of Suicides",
       color = "Sex") +
  theme_minimal()


```

# Save most recent workspace for further analyses

```{r}
# save.image(file = "./workspace.RData", safe = TRUE)  # this is calling .GlobalEnv, which is empty
save(list = ls(all.names = TRUE), file = "./workspace.RData", envir = environment())
```

# Print session info

```{r}
sessionInfo()
```